Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-24 paper/ijm.tex
Hopefully last commit before submission. Minor spelling mistakes.

There is also an increasing body of literature focusing on other languages than the local main business language. These studies generally find
that English is a significant predictor of better income
\citep{Lang2009, Casale2011, Toomet2011, Williams2011, azam+2013EDandCC, isphording2013, fabo+2017E}.
The role of other languages, in particular in multi-lingual economy, is
less clear. For instance,
\cite{Drinkwater1997} finds
a positive effect for Welsh speaking workers in
Wales and \cite{Armstrong2015} shows that bilingual native French-speakers enjoy
a significant premium both inside and outside of Quebec.
\citet{FrancoisGrin1998}, in contrast,
shows that Italian language skills in Swiss non-Italian speaking
regions are associated with a lower wage. It remains unclear if
this is a true effect or just an omitted variable bias.

@@

The literature on Central Eastern Europe (CEE) and former Soviet Union
countries is rather limited. We know that the different development
trajectories have led to disparities both favouring Russian-speaking
minority in Ukraine \citep{Constant2011} and hindering them in Estonia
\citep{Leping2008}. \citet{Kahanec2009} show that non-citizens suffer
less favourable outcomes, in particular, ethnic Russians in Estonia and
Latvia. 
\cite{Alan2015} show that Russian language proficiency is valuable in
the former USSR republic in the Caucasus (Armenia, Azerbaijan and
Georgia) by increasing employment likelihood between six and nine
percentage points. \citet{Toomet2011}, however, fails to find any
income premium from the majority language fluency in Estonia and
Latvia. This outcome is corroborated by \citet{Lindemann2013} who shows that Estonian fluency
is associated with faster job finding rate while it does not
explain the lower occupational prestige of Russian speakers. 
A number of these studies have to overcome severe data limitations,
such as lack of fine-grained information on ethnic background in
EU-wide surveys \citep{Kahanec2010}.

@@

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s. 
\Cref{fig:gdp_growth} shows that grew averaged at approximately 7\%
annually
before the financial crisis in 2008, afterwards the figure has been around 3\%.
As a by-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of transition.

@@

The economic and political reforms led to substantial
disparities in income and education opportunities
in the early 1990s \citep{Leping2008,
	lindemann+saar2011Russian2ndGeneration}. The inter-ethnic contacts
are largely limited to superficial encounters at work and in public
space, the number of close private ties remain limited
\citep{korts2009JofBaltStud}. The
separate worlds are also reflected in media which may present quite
different and occasionally antagonistic viewpoints depending on the
language \citep{Korts2002}.

@@

The language fluency variables clearly show that ethnic identity is
closely related to language use: more than 99\% of Estonians report
being fluent in Estonian, the figure for Russians is similar. We
also see that Estonians are better at English than Russians and they
are also better at Russian than vice versa.

Average age reflects the trend of younger generations
being more fluent in English and Estonian while older Estonians being
better in Russian. The family status--related variables correspond to the
respective age distribution. Not surprisingly, inter-ethnic
households include a disproportionally large share of Russian speaking
Estonians and Estonian speaking Russians.
The average experience follows the same pattern as age.
Public sector jobs are associated with disproportionally large share of
Estonian fluency among Russian workers, probably reflecting the formal
language requirements and little room for ethnic segregation in that
sector.

@@

We describe the outcome (log wage and unemployment status) $y$ of individual $i$ at year $t$ as
\begin{equation}
	\label{eq:specification}
	\begin{split}                            
		y_{it} = &\: \vec{\alpha}' \vec{X}_{it} + \vec{\beta}{}' \vec{L}_{it} +
		\eta_{i} +
		\upsilon_{t} + \epsilon_{it},                                                                             
	\end{split}
\end{equation}
where $\vec{L}$ is the vector of
language skill descriptors, $\vec{X}$ are the other individual
characteristics, $\eta_{i}$ is the individual-specific effect, $\upsilon_{t}$ are the year dummies and $\epsilon_{it}$
is the idiosyncratic error term. 
As we use panel data, we cluster the
standard errors on individuals. When analysing the long-term time
trend, we use the same specification (without year fixed effects) to
estimate separate models for each individual year.

@@

Neither Russian nor English show a comparably strong effect.
Knowledge of Russian is not associated with less unemployment for
Estonian men, and is even related to a slightly higher unemployment
for Estonian women. The latter outcome may be related to omitted
variable bias, Russian knowledge was widespread in the Soviet era and may
not have been closely associated with favourable unobserved
characteristics, however, it has been increasingly associated with age
through the period of analysis.
English skills are related to less unemployment both for Estonian men
and women, but the effect is small.
For Russians, the most important language is clearly Estonian. Those who are
fluent in that language are approximately 5 percentage point less likely
unemployed, the effect changes only a little when introducing the
individual controls. In contrast, the effect of English
is close to zero and not statistically significant. As in the
case of Estonian, individual characteristics change the estimates only
a little.

@@

English language fluency is associated with a very large
wage premium, up to 34 log points, for all the analysed groups. The number falls substantially when introducing the
individual characteristics into the model (model 2), additional
workplace characteristics have less effect. However, in all specifications it remains noticeably more
important than Russian for Estonians in all
specifications. The estimates are rather
similar for Estonian and Russian men but little lower for Russian
women than for Estonian women.

@@

As the Estonian economy went through turbulent times and rapid changes
during the 1990s, it is instructive to analyse the effect long-term
trends in the estimated effect size. Here we look at the yearly estimates from 1992 to 2012.
The reader should keep in mind that while the pre-2000 waves include retrospective income, they do not include retrospective language skills. In this way, the results before
1995 are partly based on extrapolation, assuming an individual's language
skills did not vary before that date.

@@

The effect of language fluency on unemployment is shown on
Figure~\ref{fig:long-run_unemployment}, and on wage on
Figure~\ref{fig:long-run_wage}.
The results reveal a few interesting facts. First, the effects are
relatively stable over time, indicating that neither economic
environment nor the data quality have changed rapidly. Estonian
fluency has been associated with less unemployment for both Russian men and
women since around the year 2000, in the 1990s it had little effect. Russian fluency,
in contrast, had never had an impact on Estonians' unemployment probability.

@@

For Estonian men, fluency in English is associated with less unemployment through the whole
period of observation whereas fluency in Russian has no effect.
Only a few of the results are point wise statistically significant though.
For Estonian women, both languages are associated with less
unemployment in the 1990s but not later. The picture is different for
Russians. For men, we cannot see any substantial effect of fluency in English
whereas fluency in Estonian is clearly associated with less unemployment from
the late 1990s on. For women, the picture is mostly similar, just the
effect of Estonian language fluency seems to be somewhat delayed compared to
that for men.

@@

The ethnic segregation seems to be less of an issue for women.
Unfortunately, neither \citet{Toomet2011} nor
\citet{leppik+vihalemm2015JofBaltStud} analyse men and women
separately. We find some evidence that female labor market is less
segregated along ethnic lines, and women are more frequently employed in jobs in which they need Estonian
language more, such as working in a direct contact with customers.
We also find evidence that over time Estonian is becoming more
important in the female labour market. Unfortunately, the current data
does not let us assess whether is due to decreasing segregation, or
economy--wide rise of the importance of Estonian language. No such trend is visible for
men suggesting that 20 years ---almost a generation--- after
establishing a nation state and market economy, the
labour market integration is still not complete.

@@

%\listoffixmes

>>>>

There is also an increasing body of literature focusing on other languages than the local main business language. These studies generally find
that English is a significant predictor of better income
\citep{Lang2009, Casale2011, Toomet2011, Williams2011, azam+2013EDandCC, isphording2013, fabo+2017E}.
The role of other languages, in particular in a multi-lingual economy, is
less clear. For instance,
\cite{Drinkwater1997} finds
a positive effect for Welsh speaking workers in
Wales and \cite{Armstrong2015} shows that bilingual native French-speakers enjoy
a significant premium both inside and outside of Quebec.
\citet{FrancoisGrin1998}, in contrast,
shows that Italian language skills in Swiss non-Italian speaking
regions are associated with a lower wage. It remains unclear if
this is a true effect or just an omitted variable bias.

@@

The literature on Central Eastern Europe (CEE) and former Soviet Union
countries is rather limited. We know that the different development
trajectories have led to disparities both favouring the Russian-speaking
minority in Ukraine \citep{Constant2011} and hindering them in Estonia
\citep{Leping2008}. \citet{Kahanec2009} show that non-citizens suffer
less favourable outcomes, in particular, ethnic Russians in Estonia and
Latvia. 
\cite{Alan2015} show that Russian language proficiency is valuable in
the former USSR republic in the Caucasus (Armenia, Azerbaijan and
Georgia) by increasing employment likelihood between six and nine
percentage points. \citet{Toomet2011}, however, fails to find any
income premium from the majority language fluency in Estonia and
Latvia. This outcome is corroborated by \citet{Lindemann2013} who shows that Estonian fluency
is associated with faster job finding rate while it does not
explain the lower occupational prestige of Russian speakers. 
A number of these studies have to overcome severe data limitations,
such as lack of fine-grained information on ethnic background in
EU-wide surveys \citep{Kahanec2010}.

@@

The Estonian economy, a part of the Soviet economy up to 1991, successfully
re-oriented to the Western markets and has been growing rapidly since
the mid-1990s. 
\Cref{fig:gdp_growth} shows that grew averaged at approximately 7\%
annually
before the financial crisis in 2008, afterwards the figure has been around 3\%.
As a by-product of the re-adjustment, English skills become extremely
valuable while the supply was still rather limited during the early
years of the transition.

@@

The economic and political reforms led to substantial
disparities in income and education opportunities
in the early 1990s \citep{Leping2008,
	lindemann+saar2011Russian2ndGeneration}. Inter-ethnic contacts
are largely limited to superficial encounters at work and in public
space, the number of close private ties remain limited
\citep{korts2009JofBaltStud}. The
separate worlds are also reflected in media which may present quite
different and occasionally antagonistic viewpoints depending on the
language \citep{Korts2002}.

@@

The language fluency variables clearly show that ethnic identity is
closely related to language use: more than 99\% of Estonians report
being fluent in Estonian, the figure for Russians is similar. We
also see that Estonians are better at English than Russians, and they
are also better at Russian than vice versa.

Average age reflects the trend of younger generations
being more fluent in English and Estonian while older Estonians being
better in Russian. The family status--related variables correspond to the
respective age distribution. Not surprisingly, inter-ethnic
households include a disproportionally large share of Russian-speaking
Estonians and Estonian-speaking Russians.
The average experience follows the same pattern as age.
Public sector jobs are associated with a disproportionally large share of
Estonian fluency among Russian workers, probably reflecting the formal
language requirements and little room for ethnic segregation in that
sector.

@@

We describe the outcome (log wage and unemployment status) $y$ of individual $i$ at year $t$ as
\begin{equation}
	\label{eq:specification}
	\begin{split}                            
		y_{it} = &\: \vec{\alpha}' \vec{X}_{it} + \vec{\beta}{}' \vec{L}_{it} +
		\eta_{i} +
		\upsilon_{t} + \epsilon_{it},                                                                             
	\end{split}
\end{equation}
where $\vec{L}$ is the vector of
language skill descriptors. $\vec{X}$ are the other individual
characteristics. $\eta_{i}$ is the individual-specific effect, $\upsilon_{t}$ are the year dummies and $\epsilon_{it}$
is the idiosyncratic error term. 
As we use panel data, we cluster the
standard errors on individuals. When analysing the long-term time
trend, we use the same specification (without year fixed effects) to
estimate separate models for each individual year.

@@

Neither Russian nor English language skills show a comparably strong effect.
Knowledge of Russian is not associated with less unemployment for
Estonian men, and is even related to a slightly higher unemployment
for Estonian women. The latter outcome may be related to omitted
variable bias, Russian knowledge was widespread in the Soviet era and may
not have been closely associated with favourable unobserved
characteristics, however, it has been increasingly associated with age
through the period of analysis.
English skills are related to less unemployment both for Estonian men
and women, but the effect is small.
For Russians, the most important language is clearly Estonian. Those who are
fluent in that language are approximately 5 percentage point less likely
unemployed, the effect changes only a little when introducing the
individual controls. In contrast, the effect of English
is close to zero and not statistically significant. As in the
case of Estonian, individual characteristics change the estimates only
a little.

@@

English language fluency is associated with a very large
wage premium, up to 34 log points, for all the analysed groups. The number falls substantially when introducing the
individual characteristics into the model (model 2), additional
workplace characteristics have less effect. However, it remains noticeably more
important than Russian for Estonians in all
specifications. The estimates are rather
similar for Estonian and Russian men but little lower for Russian
women than for Estonian women.

@@

As the Estonian economy went through turbulent times and rapid changes
during the 1990s, it is instructive to analyse the long-term
trends in the estimated effect size. Here we look at the yearly estimates from 1992 to 2012.
The reader should keep in mind that while the pre-2000 waves include retrospective income, they do not include retrospective language skills. In this way, the results before
1995 are partly based on extrapolation, assuming an individual's language
skills did not vary before that date.

@@

The effect of language fluency on unemployment is shown on
Figure~\ref{fig:long-run_unemployment}, and on wage on
Figure~\ref{fig:long-run_wage}.
The results reveal a few interesting facts. First, the effects are
relatively stable over time, indicating that neither the economic
environment nor the data quality have changed rapidly. Estonian language
fluency has been associated with less unemployment for both Russian men and
women since around the year 2000, in the 1990s it had little effect. Russian language fluency,
in contrast, had never had an impact on Estonians' unemployment probability.

@@

For Estonian men, fluency in English is associated with less unemployment through the whole
period of observation whereas fluency in Russian has no effect.
Only a few of the results are pointwise statistically significant though.
For Estonian women, both languages are associated with less
unemployment in the 1990s but not later. The picture is different for
Russians. For men, we cannot see any substantial effect of fluency in English
whereas fluency in Estonian is clearly associated with less unemployment from
the late 1990s on. For women, the picture is mostly similar, just the
effect of Estonian language fluency seems to be somewhat delayed compared to
that for men.

@@

The ethnic segregation seems to be less of an issue for women.
Unfortunately, neither \citet{Toomet2011} nor
\citet{leppik+vihalemm2015JofBaltStud} analyse men and women
separately. We find some evidence that female labour market is less
segregated along ethnic lines, and women are more frequently employed in jobs in which they need Estonian
language more, such as working in a direct contact with customers.
We also find evidence that over time Estonian is becoming more
important in the female labour market. Unfortunately, the current data
does not let us assess whether is due to decreasing segregation, or
economy--wide rise of the importance of Estonian language. No such trend is visible for
men suggesting that 20 years ---almost a generation--- after
establishing a nation state and market economy, the
labour market integration is still not complete.

@@
